Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2025-02-03 Manuscript_Initial.qmd
Completed first-pass revisions for Ecography

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we recorded all covariates considered for each model in our sample, including those not used in final models. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of studies considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs. We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies). Often these are included as simple covariates, as in @duggan2011's models using landcsape connectivety metrics as covariates on colonisation and extinction. Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011]. Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species. This is observed frequently in the Spotted owl literature, where Barred Owl presence is often considered as a covariate driving Spotted owl occupancy dynamics [@olson2005; ; @sovern2014; @dugger2016]. The use of these types of covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.

>>>>

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we recorded all covariates considered for each model in our sample, including those not used in final models. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of studies considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs. We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies). Often these are included as simple covariates, as in @duggan2011's models using landscape connectivity metrics as covariates on colonisation and extinction. Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011]. Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species. This is observed frequently in the Spotted owl literature, where Barred Owl presence is often considered as a covariate driving Spotted owl occupancy dynamics [@olson2005; ; @sovern2014; @dugger2016]. The use of these types of covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
